---
title: "An introduction to biodbHmdb"
author: "Pierrick Roger"
date: "`r BiocStyle::doc_date()`"
package: "`r BiocStyle::pkg_ver('biodbHmdb')`"
vignette: |
  %\VignetteIndexEntry{Introduction to the biodbHmdb package.}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
  BiocStyle::html_document:
    toc: yes
    toc_depth: 4
    toc_float:
      collapsed: false
  BiocStyle::pdf_document: default
bibliography: references.bib
---

# Purpose

biodbHmdb is a *biodb* extension package that implements a connector to
HMDB Metabolites.

We present here the different ways to search for *HMDB* [@wishart2013_HMDB]
entries with this package.

Note that the whole *HMDB* is downloaded locally by *biodb* and stored on disk,
since this is the only way to access *HMDB* programmatically.
Any search on *HMDB* is hence currently run on the local machine.

# Initialization

The first step in using *biodbHmdb*, is to create an instance of the biodb
class `Biodb` from the main *biodb* package. This is done by calling the
constructor of the class:
```{r, results='hide'}
mybiodb <- biodb::Biodb()
```
During this step the configuration is set up, the cache system is initialized
and extension packages are loaded.

We will see at the end of this vignette that the *biodb* instance needs to be
terminated with a call to the `terminate()` method.

# Creating a connector to HMDB Metabolites

In *biodb* the connection to a database is handled by a connector instance that
you can get from the factory.
biodbHmdb implements a connector to a remote database.
Here is the code to instantiate a connector:
```{r}
conn <- mybiodb$getFactory()$createConn('hmdb.metabolites')
```

# Accessing entries

To get the number of entries stored inside the database, run:
```{r}
conn$getNbEntries()
```

To get some of the first entry IDs (accession numbers) from the database, run:
```{r}
ids <- conn$getEntryIds(2)
ids
```

To retrieve entries, use:
```{r}
entries <- conn$getEntry(ids)
entries
```

To convert a list of entries into a dataframe, run:
```{r}
x <- mybiodb$entriesToDataframe(entries)
x
```

# Searching by name

We use here the generic *biodb* method `searchForEntries()` to search for
entries by name:
```{r}
id <- conn$searchForEntries(list(name='1-Methylhistidine'), max.results=1)
id
```
We limit the search result to one entry with the `max.results` field.

The first parameter is the filtering criterion, expressed as a list whose
single key (in our case) is the *biodb* field on which we want to filter.
The value is the text we want to search for.
See the documentation of `searchForEntries()` inside `?biodb::BiodbConn`.

We could also use several strings to search for, in which case an entry will be
matched if its field value contains all the specified strings:
```{r}
conn$searchForEntries(list(name=c('propanoic', 'acid')), max.results=1)
```

To look at the values of the entry, you may convert it to a data frame:
```{r}
entryDf <- conn$getEntry(id)$getFieldsAsDataframe(fields=c('accession', 'name'))
```
See table \@ref(tab:entryByNameTable) for the content of this data frame.

```{r entryByNameTable, echo=FALSE, results='asis'}
knitr::kable(entryDf, "pipe", caption="The entry returned when searching by name.")
```

# Searching inside the "description" field

Searching inside the `description` field can be done in the same way as for the
`name` field.
Here is a search with multiple strings to match:
```{r}
id <- conn$searchForEntries(list(description=c('Parkinson', 'sclerosis')), max.results=1)
id
```

Again, you can look at the values of the entry through a data frame:
```{r}
entryDf <- conn$getEntry(id)$getFieldsAsDataframe(fields=c('accession', 'name', 'description'))
```
See table \@ref(tab:entryByDescTable) for the content of this data frame.

```{r entryByDescTable, echo=FALSE, results='asis'}
knitr::kable(entryDf, "pipe", caption="The entry returned when searching by description.")
```

# Closing biodb instance

When done with your *biodb* instance you have to terminate it, in order to
ensure release of resources (file handles, database connection, etc):
```{r}
mybiodb$terminate()
```
